---
title: "Lecture 4"
author: Julia Schedler
format: revealjs
slide-number: true
scrollable: true
filters: 
  - timer
---

Pronlem 3.1 structural model for johnson and johnson

The salmon

Exponential smoothing somewhere???

## Activity 1: Export Price of Salmon (Example 3.1)
::: columns
::: {.column width="60%"}

```{r}
#| label: activity-1-template
library(astsa)

summary(fit <- lm(salmon~time(salmon), na.action=NULL))
tsplot(salmon, col=4, ylab="USD per KG", main="Salmon Export Price")
abline(fit)

# astsa version 1.5 has a new trend script that can produce Figure 3.1 with 95% CIs as follows:
trend(salmon, ylab="USD per KG", main="Salmon Export Price")
```
:::
1.  Does this time series appear stationary?

2.  The (mathematical) equation in the book for the trend line above is:

    $$
    x_t = \beta_0 + \beta_1z_t + w_t, z_t = 2003\frac{8}{12}, 2001\frac{8}{12}, \dots, 2017\frac{5}{12}
    $$

    -   There is a typo in this equation. Correct the typo. (hint: examine to the first few entries of `time(salmon)`

        ```{r}
        head(time(salmon))
        time(salmon)[1:6]
        ```

        **The typo is the second year, it should be** $2003\frac{9}{12}$

    -   Why are there fractions of the year? Explain what the fractional values mean and describe how they appear in the data set within R.

        **Note that we are dividing by 12, so the fractions represent the months. Based on the values of `time(salmon)`, January corresponds to** $\frac{0}{12}$**, so for example September is represented as** $\frac{9}{12}$**.**

3.  Interpret the estimate of the slope.

         
**Each year, the expected export price of Norwegian salmon increases by 0.2592 USD per kg.**

::: {.column width="10%"}
::: {#DecomposeJJ .timer seconds="300" starton="interaction"}
:::
:::
:::